Our Members

### 2021-08-13 pnsac/newsletter/article_EditorsNotes.tex
Changes for issue 17-1

In this issue we are recognizing the completion of a very important step in the
restoration of the North Star, namely, its Merlin engines, and our members who
have volunteered with CASM to carry out the necessary work.  This is a very
informative issue and I am sure we will all learn a great deal. I was not
aware, until putting this edition together, that the Merlin designation was in
keeping with the custom of Rolls Royce to name its aircraft engines after birds
of prey.

The article, "Four Merlins" provides the reader with a detailed history of the
restoration of these engines.   Due to its length, the article will be divided
into two editions;  starting with engines number one and two in this edition;
and three and four in the next edition. We are fortunate to have members who
have had  lengthy involvement with this part of the project. Sadly one of those
members passed away recently. Inside you will find a tribute to Tim Timmins by
the President of our Association.

We look forward to receiving your comments and suggestions regarding this
edition and for future editions whether good, bad or ugly. Please use my email
address located on the back page.

Good reading.

>>>>

Unfortunately COVID-19 and its restrictions continue to hamper the restoration
of the North Star. So it's a good time to look at what has been accomplished to
date by the restoration team which consists of both Museum staff and volunteers.
This edition will continue the focus on the Merlin engines. As many of you know
many of the volunteers are also members of our organization.

You will find photos of some of those volunteers in the technical article on
engines three and four. You will also find a interview with Garry Dupont who has
had a leading role in the engine shop work.

In the next edition we expect to have more articles about our members
experiences in the restoration project.

In the meantime if you have any comments on this edition or suggestions for
future editions please send them to me.

### 2021-08-13 pnsac/newsletter/article_Feature1.tex
Changes for issue 17-1

\title{North Star Propeller Restorations}
%\subtitle{Part 4}
\author{Harry Hope, with assistance from Michael Hope.}

@@

\textit{Hope Aero has played a major role in the restoration of the North Star.
The following article provides some background about the company and its work
on the North Star’s propellers. }\\

Hope Aero was formally government approved to certify aircraft propellers in
1992, at their facility in Mississauga, Ontario. Harry Hope however, already
had 40 years of propeller overhaul experience on many varieties that included
Hamilton Standard propellers, which were quite common, as well as Hartzell,
McCauley, and Sensenich. Harry's first restoration to flying condition, was the
propeller for the Lysander, which was restored by the Air Force in Winnipeg for
their Centennial project in 1967. This Lysander is now displayed at CASM. Since
that time, they have restored several propellers, some to flying condition,
three are the Lancaster for the Canadian Warplane Heritage, Super Constellation
(Super Star) for Lufthansa, and the Honningstad C5 Polar for Norway which is on
display at the Aviation museum in Boda, Norway.

 Since that time, their capabilities have added– MT, Dowty, Hoffman, and
Hamilton Standard Dash 7 and 8, and later the Dowty propeller on the Q400.  As
their customer base expanded to include different types of aircraft, the
requirement for them to service more types became evident. They kept up with
the required technology.

Along the way, they also expanded into aircraft Wheel and Brake overhaul,
sales, service and on wing dynamic balancing of fixed wing and Helicopters.
Harry retired in 2001, but the third generation is now capably looking after
the business.

In the spring of 2006, Hope Aero took on the task of restoring the four North
Star propellers, for the "North Star Restoration" project. The first challenge
was to remove the propellers from the engines as neither the restoration crew
nor Hope Aero had any recent experience removing this type of propeller.
Michael Hope made the trip to Ottawa to assist with this operation and to help
prevent any damage, which turned out to be exceedingly difficult due to the
corrosion, starting with the spinner and then the propeller. Once the propeller
was removed the blades were removed for shipping to Hope Aero.

The first 43D50 Hamilton Standard Hydromatic propeller was received at Hope
Aero in 2006. It was completely disassembled, cleaned and inspected. The
external parts were very badly corroded, with most damage being to the aluminum
blades and dome shells, and rust to the steel hub and tail shaft, which was
likely caused by the outside storage environment. The inside parts were in
relatively good condition. It was established, that both technical and visual
effort would be required to restore the propeller to an acceptable display
condition.

The restoration work commenced with the blades, the alcohol de-icer feed shoes,
troughs and paint being removed. The corrosion of the blade surfaces was such
that removal by grinding was abandoned, as the inner half of the blades are
shot peened and grinding would remove this and the authenticity. It was decided
to media blast the complete airfoil surfaces with an aluminum oxide media, thus
removing the surface corrosion but leaving a good surface for paint to be
adhered to; an aviation grade grey paint was applied with yellow stripes at the
tips. New alcohol feed shoes were installed, however the troughs, which direct
the alcohol to the shoes, were not available. As the troughs are hidden inside
the spinner, it was determined that they could be omitted.
The external hub, barrel halves, spider tail shaft, slinger ring, de-icer
tubes, and dome retaining nut, were extremely rusty. They were glass bead
blasted to remove the rust; this did not remove the heavy pitting but did leave
a surface that could be cadmium plated.  After plating, a clear coat material
was sprayed on these parts to improve the resistance to rust. The internal
surfaces were in good condition, and they were well oiled.

A replacement dome shell in relatively good condition was found, requiring only
to be sanded, anodized, and be reinstalled with new seals.

The propeller was then completely assembled to confirm that all the refurbished
parts fit together. The blades were then removed to facilitate being shipping
back to the museum.

All of the four propellers were  found to be in similar condition, and all were
restored in a similar manner.

>>>>

\title{The Four Merlins: The Second Two}
%\subtitle{Part 4}
\author{Based on Bruce Gemmill's notes; reviewed by Garry Dupont, and with editing assistance from Richard Lodge}

@@

% \paragraph*{Engine 3}

After its removal in the Spring of 2012, engine \#3 was completely stripped of
all accessories, pipes and hoses, and disassembled.  The block, cylinders and
pistons were cleaned, and work was begun to reassemble this engine.  It was at
this point that a lack of volunteers started to slow down the restoration
process.

The engine rebuild was nearing completion by April 2013, with magnetos and a few
other accessories left to be installed.  The basic engine frame was back
together.  The three radiator sections were cleaned and painted, then attached
to the front of the frame.  The rear cowl ring had been disassembled, and new
mounting hardware was fitted before the parts were painted and reassembled.  The
front cowl assembly (radiator cover) was also cleaned, repaired and polished,
ready to be installed.  Other cowl panels were then cleaned, repaired and
polished, ready to install once the complete engine was in place.  Work was also
underway on a host of pipes, cables, accessories, and fittings needed to
complete the assembly of the engine and the Quick Engine Change (QEC) prior to
installation on the aircraft.

By September 2013 engine \#3 was almost completely reassembled, with only the
supercharger and inter cooler preheat assemblies left to install on the engine,
which had been moved to the QEC engine frame in early summer.  The work slowed,
due to summer down time for staff and volunteers. The auxiliary gearbox was
removed from the engine firewall and was then being disassembled.  This was the
last major assembly that had to be overhauled before the engine could be placed
back on the aircraft.

Engine \#3 was nearing completion by December 2013. The supercharger and inter
cooler preheat assemblies had been installed, along with the propeller reduction
gearbox.  While this work was being carried out by the engine crew, the
remainder of the volunteer workforce busied themselves completing the numerous
cowl panels, pipes, hose and other pieces needed to complete the engine.  Two
items that seemed to take forever were the large steel exhaust shrouds.  A lot
of hammering was needed to pound out years of dents, then numerous rivets and
all the Dzus fasteners and springs needed to be replaced.  The complex shape of
the exhaust shroud made this challenging work.  Further delaying completion was
the search for suitable clear coat to protect the steel from corrosion.  The old
clear coat used successfully on engines \#1 and \#2 was no longer available.  A
suitable replacement was located, the shrouds were clear coated and serial
numbers were stenciled on.

The auxiliary gearbox was disassembled, and significant corrosion was found on
the bearings inside.  Bearings had to be ordered from England.  Meanwhile, the
gearbox was cleaned and painted.  At the same time the large DC generator was
dismantled, repainted and reassembled, as well as the air pump, and the
tachometer generator and propeller synchronizer.  Once the new bearings were
installed, the gearbox was completed, and the generator, and air pump were
attached.  Finally, the gearbox was installed on the firewall awaiting the
installation of the engine.

In April 2013 \#3 engine frame was completed after the installation of the engine
and supercharger.  Fitting the many cowl panels proved difficult, due to the
extensive rework that many of these items went through, causing some alignment
difficulties when fitting these panels onto the engine frame.  This work was
done in July 2014, and the completed engine was successfully installed on the
aircraft.

\begin{figure}[H]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{img-1-merlins-aug2021.png}
   %caption of the figure 
   \caption*{\small \em Volunteers Bill Tate and Garry Dupont discussing work on Engine 3.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:img1}
\end{figure}

% \paragraph*{Engine 4}

The last engine to undergo restoration was removed iand work
immediately began in the Engine Shop to disassemble the engine.  All cowl
panels, pipes, hoses and accessories were removed and stored until space was
available to restore these items. The engine was moved to the yellow rotating
stand at this point. The priority was to disassemble the engine.]. The crank
case was cleaned and clear coated before installing back into the yellow stand.
The refurbished crankshaft was installed in the engine block, and the connecting
rods and pistons cleaned and polished. On this engine we were able to remove the
propeller shaft from the reduction gear box, unlike engine \#3 the shaft could
not be removed.  Work was then begun on the more complex cylinder heads and
valve train assemblies.

Work then proceeded on much of the ancillary equipment for the engine. Once the
frame was completed and placed on the blue stand, the engine block was removed
from the yellow stand and installed in the frame. The supercharger \& inter
cooler were completed and installed, as well as the reduction gearbox. The
radiators and header tank were restored and painted, and two of the three
radiator sections were installed. Restoration work was then begun on the air
intake system, and many of the fuel and oil lines.  The starter motor was found
to be very corroded and disassembly and repair not viable. A replacement unit
was located and installed. The fire suppression lines were complete and would be
installed along with the cowls when ready. Some of the cowl panels were
restored, although it was found that this engine had more corrosion than the
other three engines. Possibly it had been in service longer.  This meant that
more work was required to refurbish each of the many panels.

After restarting work on the aircraft in late January 2018, the conservation
effort focused on the completion of engine number \#4 accessories and cowl
panels.   Many of the cowl panels required extensive repairs, including making
many new panel stiffeners, since the old ones were often badly corroded.
Besides painting and polishing, the refurbished panels also required new part
number and serial number stencils, and this led to investigating missing
stencils from the same panels on earlier engines.

Before final assembly could take place, the radiator cowl panel had to be fitted
which was done by December. The panels were fitted and installed with some
fasteners being replaced. The front support was installed for transportation.

\begin{figure}[H]
    \vspace{2em}
    \centering
    %name of the graphic, without the path AND in EPS format:
    \includegraphics[scale=0.5]{img-2-merlins-aug2021.png}
    %caption of the figure 
    \caption*{\small \em Engine 4 completed and ready to be moved to the Storage Hangar.}
    %label of the figure, which has to correspond to \ref{}:
    \label{fig:img2}
 \end{figure}

 \begin{figure}[H]
    \vspace{2em}
    \centering
    %name of the graphic, without the path AND in EPS format:
    \includegraphics[scale=0.5]{img-3-merlins-aug2021.png}
    %caption of the figure 
    \caption*{\small \em Some of the crew who worked on the restoration of Engine 4 (Richard Lodge, R\'{e}j Demers, Garry DuPont, Ted Devey, Charles Baril).}
    %label of the figure, which has to correspond to \ref{}:
    \label{fig:img3}
 \end{figure}

On December 19, 2019, the engine, still on its blue stand, was towed by the
Museum forklift, from the Restoration Shop to the Storage Hangar where it now
sits next to the North Star prior to re-installation on the aircraft.

\begin{figure}[H]
    \vspace{2em}
    \centering
    %name of the graphic, without the path AND in EPS format:
    \includegraphics[scale=0.5]{img-4-merlins-aug2021.png}
    %caption of the figure 
    \caption*{\small \em R\'{e}j Demers (Museum Special Projects Manager) very carefully bringing the Museum forklift to lift one end of Engine 4 ready for moving from the Engine Shop to the Storage Hangar on 13 December 2019.}
    %label of the figure, which has to correspond to \ref{}:
    \label{fig:img4}
 \end{figure}

 \begin{figure}[H]
    \vspace{2em}
    \centering
    %name of the graphic, without the path AND in EPS format:
    \includegraphics[scale=0.5]{img-5-merlins-aug2021.png}
    %caption of the figure 
    \caption*{\small \em Engine 4 starts it long, slow, cold 20 minute journey from the Engine Shop to the Storage Hanger to join the North Star aircraft.}
    %label of the figure, which has to correspond to \ref{}:
    \label{fig:img5}
 \end{figure}

 \begin{figure}[H]
    \vspace{2em}
    \centering
    %name of the graphic, without the path AND in EPS format:
    \includegraphics[scale=0.5]{img-6-merlins-aug2021.png}
    %caption of the figure 
    \caption*{\small \em The Museum’s forklift, driven with great skill by  R\'{e}j Demers, hauls Engine 4 towards the Storage Hangar.  The blue engine stand and its tiny wheels certainly was not made for racing!}
    %label of the figure, which has to correspond to \ref{}:
    \label{fig:img6}
 \end{figure}

2020 Finally early in 2020, before the lock down,  propeller \#4 was assembled
and stored on the refurbished prop stand in the Storage Hangar ready to mount on
engine \#4.

% \begin{figure}[H]
%    \vspace{2em}
%    \centering
%    %name of the graphic, without the path AND in EPS format:
%    \includegraphics[scale=0.5]{four-merlins-1.png}
%    %caption of the figure 
%    \caption*{\small \em North Star being moved inside hangar 11 August 2005 .}
%    %label of the figure, which has to correspond to \ref{}:
%    \label{fig:tim}
% \end{figure}

% \begin{figure}[htbp]
%    \vspace{2em}
%    \centering
%    %name of the graphic, without the path AND in EPS format:
%    \includegraphics[scale=0.5]{four-merlins-2.png}
%    %caption of the figure 
%    \caption*{\small \em  Mike Hope and Bruce Gemmill assembled Prop 1ready to
% install on Engine 1 October 26, 2008.}
%    %label of the figure, which has to correspond to \ref{}:
%    \label{fig:tim}
% \end{figure}

% \begin{figure}[htbp]
%    \vspace{2em}
%    \centering
%    %name of the graphic, without the path AND in EPS format:
%    \includegraphics[scale=0.55]{four-merlins-3.png}
%    %caption of the figure 
%    \caption*{\small \em Engine 1 assembled in the Conservation shop Feb 2,
% 2010.}
%    %label of the figure, which has to correspond to \ref{}:
%    \label{fig:tim}
% \end{figure}

% \begin{figure}[htbp]
%    \vspace{2em}
%    \centering
%    %name of the graphic, without the path AND in EPS format:
%    \includegraphics[scale=0.55]{four-merlins-4.png}
%    %caption of the figure 
%    \caption*{\small \em Firewall ready, auxiliary gearbox installed Jan 7,
% 2010.}
%    %label of the figure, which has to correspond to \ref{}:
%    \label{fig:tim}
% \end{figure}

% \begin{figure}[htbp]
%    \vspace{2em}
%    \centering
%    %name of the graphic, without the path AND in EPS format:
%    \includegraphics[scale=0.55]{four-merlins-5.png}
%    %caption of the figure 
%    \caption*{\small \em Engine 1 complete at the Restoration Shop Feb 9, 2010
% .}
%    %label of the figure, which has to correspond to \ref{}:
%    \label{fig:tim}
% \end{figure}

% \begin{figure}[htbp]
%    \vspace{2em}
%    \centering
%    %name of the graphic, without the path AND in EPS format:
%    \includegraphics[scale=0.45]{four-merlins-6.png}
%    %caption of the figure 
%    \caption*{\small \em Engine 1 installed Feb 24, 2010. Shown top left to
% bottom right, engine
% supervisor Ted Devey,  Bill Tate, Ron Lemieux, Giuseppe Zanetti, John Corby,
% Garnet Chapman, Volunteer Project Manager Jim Riddoch, John Tasseron, Bruce
% Gemmill, Museum Project Coordinator Mike Irvin, Charles Baril, and then
% Association President Tim Timmins.}
%    %label of the figure, which has to correspond to \ref{}:
%    \label{fig:tim}
% \end{figure}

% \paragraph*{Engine 2}\

% \begin{figure}[htbp]
%    \vspace{2em}
%    \centering
%    %name of the graphic, without the path AND in EPS format:
%    \includegraphics[scale=0.5]{four-merlins-8.png}
%    %caption of the figure 
%    \caption*{\small \em Engine 2 with cowl panels removed, April 6, 2010.}
%    %label of the figure, which has to correspond to \ref{}:
%    \label{fig:tim}
% \end{figure}

% \begin{figure}[htbp]
%    \vspace{2em}
%    \centering
%    %name of the graphic, without the path AND in EPS format:
%    \includegraphics[scale=0.5]{four-merlins-9.png}
%    %caption of the figure 
%    \caption*{\small \em Engine 2 removed April 14, 2010.}
%    %label of the figure, which has to correspond to \ref{}:
%    \label{fig:tim}
% \end{figure}

% \begin{figure}[htbp]
%    \vspace{2em}
%    \centering
%    %name of the graphic, without the path AND in EPS format:
%    \includegraphics[scale=0.5]{four-merlins-10.png}
%    %caption of the figure 
%    \caption*{\small \em Stan Rideout working to reassemble Engine 2 in engine
% shop, Oct 7, 2010.}
%    %label of the figure, which has to correspond to \ref{}:
%    \label{fig:tim}
% \end{figure}

% \begin{figure}[htbp]
%    \vspace{2em}
%    \centering
%    %name of the graphic, without the path AND in EPS format:
%    \includegraphics[scale=0.5]{four-merlins-11.png}
%    %caption of the figure 
%    \caption*{\small \em Austin (Tim) Timmins.}
%    %label of the figure, which has to correspond to \ref{}:
%    \label{fig:tim}
% \end{figure}

% \begin{figure}[htbp]
%    \vspace{2em}
%    \centering
%    %name of the graphic, without the path AND in EPS format:
%    \includegraphics[scale=0.5]{four-merlins-12.png}
%    %caption of the figure 
%    \caption*{\small \em Jim Riddoch and Bill Tate fill the radiators with
% inhibiting oil.}
%    %label of the figure, which has to correspond to \ref{}:
%    \label{fig:tim}
% \end{figure}

% \begin{figure}[htbp]
%    \vspace{2em}
%    \centering
%    %name of the graphic, without the path AND in EPS format:
%    \includegraphics[scale=0.7]{four-merlins-13.png}
%    %caption of the figure 
%    \caption*{\small \em Engine 2 on rotating assembly stand.}
%    %label of the figure, which has to correspond to \ref{}:
%    \label{fig:tim}
% \end{figure}

% \begin{figure}[htbp]
%    \vspace{2em}
%    \centering
%    %name of the graphic, without the path AND in EPS format:
%    \includegraphics[scale=0.45]{four-merlins-14.png}
%    %caption of the figure 
%    \caption*{\small \em  Engine Nr 2 being re-assembled in the engine shop.}
%    %label of the figure, which has to correspond to \ref{}:
%    \label{fig:tim}
% \end{figure}

% \begin{figure}[htbp]
%    \vspace{2em}
%    \centering
%    %name of the graphic, without the path AND in EPS format:
%    \includegraphics[scale=0.5]{four-merlins-15.png}
%    %caption of the figure 
%   % \caption*{\small \em .}
%    %label of the figure, which has to correspond to \ref{}:
%    \label{fig:tim}
% \end{figure}

### 2021-08-13 pnsac/newsletter/article_MemberInterview.tex
Changes for issue 17-1

\title{Our Members}
\author{Interview with Garry Dupont}

@@

\end{multicols}

@@

\begin{multicols}{2}

1.\textit{What is your background in aviation?}

I graduated from the Confederation College Aircraft Maintenance course
in the Spring of 1975. After returning to my home town in the Red Lake
District I gained employment with Red Lake Seaplane Service. After
completing the course I had to apprentice for 18 months before writing
the final exam for my Aircraft Maintenance Engineers' License. I
successfully wrote my final exam in December 1976 and obtained my AME
certification. In February 1977 I was hired by Perimeter Aviation in
Winnipeg Manitoba. There I obtained my Turbine Engine and
Pressurization Aircraft endorsements.  I started my 30 year career with
the RCMP Air Services Branch in February 1981. My career took me to
Edmonton Alberta, Goose Bay Labrador, Winnipeg Manitoba then finally to
Ottawa. I finished my career as Quality Assurance Manager.

2. \textit{How long have you been involved with Project North Star and how and
why did you get involved?}

I attended the very first meeting held for Project North Star. This was
at my wife's suggestion that it would be a good retirement pastime.

%\begin{figure}[htbp]
%   \vspace{2em}
%   \centering
%   %name of the graphic, without the path AND in EPS format:
%   \includegraphics[scale=0.5]{fwd_cockpit_complete.eps}
%   %caption of the figure 
%   \caption*{\small \em Forward cockpit -- complete.}
%   %label of the figure, which has to correspond to \ref{}:
%   \label{fig:fwd_cockpit_complete}
%\end{figure}

3. \textit{What has been the history of your involvement to date?}

While I was employed I worked an evening shift every 4 weeks so I would
come in to CASM in the morning for a few hours. Upon my retirement Mike
Irving had me work in the engine shop, which I have been doing since my
retirement. I also have been on the board of directors for a number of
years.

%\begin{figure}[htbp]
%	\vspace{2em}
%	\centering
%	%name of the graphic, without the path AND in EPS format:
%	\includegraphics[scale=0.9]{resources/PNS2008-10-26_23-31-42.png}
%	%caption of the figure 
%	\caption*{\small \em Propeller assembly.}
%	%label of the figure, which has to correspond to \ref{}:
%	\label{fig:propeller}
%\end{figure}

%\begin{figure}[htbp]
%	\vspace{2em}
%	\centering
%	%name of the graphic, without the path AND in EPS format:
%	\includegraphics[scale=1.3]{resources/NorthStarCockpitDec2011.png}
%	%caption of the figure 
%	\caption*{\small \em Forward cockpit -- complete.}
%	%label of the figure, which has to correspond to \ref{}:
%	\label{fig:fwd_cockpit_complete}
%\end{figure}

4. \textit{What has been the highlight of your involvement?}

The biggest highlight for me was installing number 2 engine. It was
about half completed when I started in the engine shop. This is the
largest engine I have ever installed in my entire career.

%\begin{figure}[htbp]
%   \vspace{2em}
%   \centering
%   %name of the graphic, without the path AND in EPS format:
%   \includegraphics[scale=0.75]{resources/NavigatorRack.png}
%   %caption of the figure 
%   \caption*{\small \em Navigator rack.}
%   %label of the figure, which has to correspond to \ref{}:
%   \label{fig:nav-radio-rack}
%\end{figure}

5. \textit{What has been the most challenging part of your involvement?}

One of the most challenging parts was learning how the engine worked
with all its systems. I found a RCAF training manual in the office and
took it home. Over several months of reading I started to understand
what made this monster of old technology tick. Another challenge was
learning the British Hardware system and tools. A further challenge was
reading  the old manuals which were written before they were
standardized by the ATA chapter system which all aircraft manuals
follow today. Learning the art of conservation was also a  challenge
but was made easier by the guidance of Mike Irving.

>>>>

\title{Our Members}
\author{This is the edited transcript of an interview with Garry Dupont by Richard Lodge.}

@@

\textit{Garry why did you join Project North Star?}

I joined it when I was working full time and it was a discussion between me and
my wife as I knew retirement was looming and it was what I was going to do. I
had seen the article that was in the Ottawa Citizen with Robert Holmgren and Tim
Timmins requesting people to attend a meeting starting the project, and my wife
suggested I go to it, which I did and it gave me something to look forward to.
Even when I was employed I was able to physically contribute by coming in the
odd day and helping out with special projects and overhaul a few components and
stuff like that.\\

\noindent\textit{So you worked only in the engine shop?}

No.  When initially I was working, I worked on various other things and only
after I fully retired, I showed up and said Mike Irvin "I am here. I'll be here
for four days a week".  That's when he asked me to go into the engine shop
because they needed help in there.\\

\noindent\textit{Could you have worked anywhere but you chose to work in the engine shop?}

Actually, Mike chose for me.  Mike was the one who said go to the engine shop.\\

\noindent\textit{When you started with Project North Star would you have felt
that you had most experience of aviation work on engines, or could you gone
anywhere on the plane?}

I could have gone anywhere on the plane. Of this particular engine I had no
knowledge. I had never worked on these liquid cooled engines before.  I never
worked on a piston engine of that size before and actually I hadn't worked on
piston engines for quite a number of years because with my employment most of my
airplanes were with turbines.  We had some small piston engined airplanes of
which I had limited experience.\\

\noindent\textit{So quite a bit of your work in the engine shop was a learning
experience apart from learning about how the Merlin 622 was assembled, you were
also learning a bit about actually working on a large piston engine aircraft?}

That's correct and the systems on this particular engine were completely
different from anything I had ever worked on and I came across  an RCAF training
manual, which somebody had donated and I took it home and read it from one end
to the other over a couple of weeks. I was able to learn a considerable amount
about the engine and understand the systems and it started making sense to me,
so when you start assembling and disassembling and you look at a line you know
what that the line is for. You know what the electrical connections are for. You
know what system they belong to so it makes that part of the process much
easier.\\

\noindent\textit{I know from when we first started working together, you said
you worked for Perimeter Airlines in Winnipeg.  When you were working for
Perimeter, you were working mostly on Metroliners. Correct?}

Yes.  We had 2 Metroliners when I was there, but we also had 4 piston operated
Beechcraft Queenairs, we had 6 other Beechcraft.\\

\noindent\textit{When you worked at Perimeter, were you working on any part of
the plane and you weren't specialising in any particular part, so moving into
solely engines was really a new area of work for you?}

That's correct.\\

\noindent\textit{When you first started working in the engine shop, who were you
working with?}

There was Ted Devey, John Tasseron, Michel Lacasse and two other
fellows, the names escape me.\\

\noindent\textit{I asked a few minutes ago.  If you were only interested in
working on the engines.  Where else would you like to work on the plane? So you worked only in the engine shop?}

Anything actually, I don't do well with sheet metal work, I have done some,
minor bits. I could develop those skills a bit better.  Hydraulic systems; I
have a fairly good interest in, landing gear, the flaps, the hydraulics and all
those systems.  I have learned where the pumps are and how all that works now.
I have also read up little bit on the landing gear to understand it a bit better
and how it works and that would be interesting.\\

\noindent\textit{So you could, if they had been doing the landing gear at the
time you started have been quite interested in working on that side of things?}

Yes.\\

\noindent\textit{What would you say in retrospect was the most difficult part of
the work you were having to do on engines 3 and 4?  Was it lack of paperwork,
was it corrosion or what was it?}

The most difficult part was dealing with corrosion because of the amount of
corrosion in the engine and also the decision-making between my career and what
I would do and what the museum wanted in restoration. The hardest learning part
was learning about restoration and how to do it properly. I have had some good
mentors.  and I have learned fairly well on what is required to do and what to
look forward to in the future when planning the project; how to tackle it from a
restoration perspective rather than a maintenance perspective.\\

\noindent\textit{Was Mike Irvin one of your mentors helping you to know the
various procedures needed for restoration?}

Yes, Mike was the main mentor.  He was the one that took me under his wing and
showed me the requirements and the basis of doing restoration.\\

\noindent\textit{When you started working there, you weren't really in charge of
the engine because it was a combination of you and Ted Devey who was still
working there and who had worked on engines 1 and 2.  So you were working with
him under Mike Irvin?}

Yes.  That's right.\\

\noindent\textit{You feel one of the most difficult parts was to make the shift
from keeping an aircraft flying to restoring and conserving everything?}

That's right, yes.\\

\noindent\textit{Having done that and got into that sort of mindset, what would
you feel was the most difficult part of doing the engine, of restoring the
engines that you had to be involved with, or you had to work with.  Which part,
I am thinking of?}

The biggest part with the engines was the cleaning.  Believe it or not, the
engines and the old oil had been sitting there for 35 years, it was very
congealed and very difficult to remove.  Cleaning with solvents is not a
pleasant thing to do and that was about the only method we could use was to
clean and put it in the parts washer.  Once you put a part in the parts washer
you also had to be very careful because if it was steel, when it was removed
from the part washer it would flash freeze almost immediately, so it had to be
removed and treated straightaway with some sort of preserving oil to prevent the
surface from oxidizing.\\

\noindent\textit{I very much remember John Tasseron working on the pistons and
all work of getting the pistons out of the block  and then John working on the
cleaning the piston rings.  Were you very much involved in that side of things?}

Yes.  Cleaning up the piston rings was a real tedious job where you could spend
a whole week on one piston ring trying to free it up by scraping away the
corrosion, by tapping it to free the rings.  Sadly, a lot of them broke.  There
was nothing we could do to free them up.  We tried various chemicals, soaking
procedures, vibration procedures, putting them in sonic cleaners so that the
vibration would free up the dirt, but nothing seemed to work really well, except
getting in there with a little scraper and trying to scrape them clean.\\

\noindent\textit{And when you did manage to free the parts and get them clean,
you couldn't use a glass bead machine to do any cleaning because the parts were
aluminum so we had to use Scotch Brite and other things to clean them?}

Correct.  The purpose of glass beading is to remove corrosion and not to clean.
If you glass bead a surface, it changes the outer texture of it considerably. In
the engine shop we had at Perimeter, the only time they used glass beading was
for removing the outside surface of the engine block and then they could prepare
it for painting.\\

\noindent\textit{Now that you have finished the main part of the work on engines
3 and 4, what are the main additional parts or work still to be done on engines
3 and 4 and maybe on engines 1 and 2?}

R\'{e}j [Demers] wants to remove the engines so we will reinstall engine 4 when
it is ready and then we will go back and will work on each engine successively
and then work on the firewall and behind the firewall with the engine removed.
At the time when we do this we will inspect the engines to see if corrosion has
come back in some areas and I notice on some of the engines during Canada Day
when we removed parts and cowlings to display the engine that there was
corrosion showing up in certain areas where it had not been treated properly.
So when we remove the engines, and R\'{e}j agrees with me, we will put them back
in the engine shop and go over them thoroughly.  We have also found some issues
with some of the cowl parts just not being adjusted properly.  So in the end, I
would like to have all four engines installed on the wings properly treated so
that we don't have to worry about them again and also the cowls adjusted and
done properly so that at any time we can remove the cowl parts and reinstall
them with ease.  So that in the future, if you want to display the engine again,
you could effectively remove all the cowls on all four engines with ease and
then reinstall them without making any big issue of it.\\

\noindent\textit{Effectively you have got to remove any distortions in the cowl
panels. So those are the things to be done?  When you are finished working on
the Merlin engines, what would you like to do?  This is purely a personal thing
and nothing to do with the Museum.  More engines on a different plane or working
on the airframe?}

It really doesn't matter to me, whatever the Museum sees fit as a project, I am
willing to tackle it.

% \end{multicols}

@@

% \begin{multicols}{2}

%1.\textit{What is your background in aviation?}
%
%
%2. \textit{How long have you been involved with Project North Star and how and
%why did you get involved?}
%
%3. \textit{What has been the history of your involvement to date?}
%
%4. \textit{What has been the highlight of your involvement?}
%
%5. \textit{What has been the most challenging part of your involvement?}
